### 2016-11-14 Manuscript/SW_early_life_telomeres.Rmd
Home edits

Including birth year as a random effect in a LMM of RTL and (log) age resulted in a significantly improved model to one with only individual ID as a random effect (P = `r round3(anova(RTL_LogAge,RTL_LogAge_LY)$"Pr(>Chisq)"[2])`). Allowing the slope of the age effect to vary among birth years resulted in a slightly significant improvement on one in which only the intercept was allowed to vary (P = `r round3(anova(RTL_LogAge_LY,RTL_LogAge_AgeLY)$"Pr(>Chisq)"[2])`). However, a model including catch year and individual ID as random effects was significantly better than a model indluding birth year and individual ID (P = `r round3(anova(RTL_LogAge_LY,RTL_LogAge_CY)$"Pr(>Chisq)"[2])`), and a model that allowed the slope of the age effect to vary among catch years was better still (P = `r round3(anova(RTL_LogAge_CY,RTL_LogAge_AgeCY)$"Pr(>Chisq)"[2])`).

Longitudinal data showed that early-life RTL was weakly, but significantly, related to adult RTL (R^2^ = `r round3(summary(lm(RTL1~RTL,data = dd3))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,3])`; P = `r round3(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,4])`; Fig. 1B). As with cross-sectional RTL, $\Delta$RTL decreased with age (R^2^ = `r round3(summary(lm(DeltaRTL~LogAge,data = dd3))$r.squared)`; t = `r round2(summary(lm(DeltaRTL~LogAge,data = dd3))$coefficients[2,3])`; P = `r round3(summary(lm(DeltaRTL~LogAge,data = dd3))$coefficients[2,4])`), suggesting that telomere shortening was highest in early life. Although both or cross-sectional and longitudinal data indicated a general trend of telomere shortening, we found that `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our $\Delta$RTL measurments were positive, and thus indicative of telomere lengthening. To test whether the apparent telomere lengthening in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual (see methods for details). We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig. 1C). Importantly, splitting the longitudinal data into instances of shortening ($\Delta$RTL < 0) and lenghtning ($\Delta$RTL > 0) revealed that not only did we observe significantly more shortening within individuals compared to within samples (Wilcoxon test: W = `r round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$statistic,0)`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also significantly more lengthening (W = `r as.character(round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$statistic,0))`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`).

#####Spatiotemporal variation in telomere dynamics and the environment
A linear mixed model of RTL in relation to social and ecological environmental variables revealed age, tarsus length, sex and annual variation in food availability were significantly related to RTL (Table 1). Model selection revealed that sex only appeared in models where tarsus length was also present (Table S1), and sex was no longer significant when tarsus length was removed from the full model (P = ). RTL increased with tarsus length in males, but decreased in females. In general, the model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-average confidence intervals not overlapping zero (Table S1; Fig. S1).

$\Delta$RTL did not differ significantly among cohorts (F = `r round2(anova(lm(DeltaRTL~factor(FieldPeriodID),data = dd3))[1,4])`; P = `r round(anova(lm(DeltaRTL~factor(FieldPeriodID),data = dd3))[1,5])`), although power for this analysis was limited as longitudinal sample sizes within cohorts were low. Model selection revealed that the top model explaining $\Delta$RTL was the null model (Table S2).

>>>>

Including birth year as a random effect in a LMM of RTL and (log) age resulted in a significantly improved model to one with only individual ID as a random effect (P `r round3(anova(RTL_LogAge,RTL_LogAge_LY)$"Pr(>Chisq)"[2])`; Fig. 2A). Allowing the slope of the age effect to vary among birth years resulted in a slightly significant improvement on one in which only the intercept was allowed to vary (P = `r round3(anova(RTL_LogAge_LY,RTL_LogAge_AgeLY)$"Pr(>Chisq)"[2])`; Fig. 2B). However, a model including catch year and individual ID as random effects was significantly better than a model indluding birth year and individual ID (P `r round3(anova(RTL_LogAge_LY,RTL_LogAge_CY)$"Pr(>Chisq)"[2])`) suggesting that variation among catch years was greater than among cohorts (Fig. 2C), and a model that allowed the slope of the age effect to vary among catch years was better still (P = `r round3(anova(RTL_LogAge_CY,RTL_LogAge_AgeCY)$"Pr(>Chisq)"[2])`; Fig. 2D). Using longitudinal data and $\Delta$Age as an explanatory variable we found a similar pattern - a model including catch year as a random effect was better than that did not allow for temporal variation in RTL (P = `r round3(anova(RTL_DeltaAge,RTL_DeltaAge_CY)$"Pr(>Chisq)"[2])`), and better than one with birth year as a random effect (P `r round3(anova(RTL_DeltaAge_LY,RTL_DeltaAge_CY)$"Pr(>Chisq)"[2])`). However a model allowing the slope of the $\Delta$Age effect was not better than one in which only the intercept varied (P = `r round3(anova(RTL_DeltaAge_CY,RTL_DeltaAge_AgeCY)$"Pr(>Chisq)"[2])`).

Within individuals, longitudinal data showed that RTL measurements taken from different timepoints were weakly, but significantly, positvely correlated (R^2^ = `r round3(summary(lm(RTL1~RTL,data = dd3))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,3])`; P = `r round3(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,4])`; Fig. 1B). Although both or cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our $\Delta$RTL measurments were positive, and thus indicative of telomere lengthening. To test whether the apparent telomere lengthening in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual (see methods for details). We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig. 1C). Importantly, splitting the longitudinal data into instances of shortening ($\Delta$RTL < 0) and lenghtning ($\Delta$RTL > 0) revealed that not only did we observe significantly more shortening within individuals compared to within samples (Wilcoxon test: W = `r round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$statistic,0)`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also significantly more lengthening (W = `r as.character(round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$statistic,0))`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`).

#####Spatiotemporal variation in telomere dynamics and the environment
A linear mixed model of RTL in relation to social and ecological environmental variables revealed age, tarsus length, sex and annual variation in food availability were significantly related to RTL (Table 1). Model selection revealed that sex only appeared in models where tarsus length was also present (Table S1), and sex was no longer significant when tarsus length was removed from the full model (estimate = `r round3(fixef(RTL_Full_NoTarsus)['SexMales'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_NoTarsus)['SexMales',],lessthan = F),collapse = ', ')`). RTL increased with tarsus length in males, but decreased in females. The model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-averaged confidence intervals not overlapping zero (Table S1; Fig. S1).

Model selection revealed that the top model explaining $\Delta$RTL was...

### 2016-11-14 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Home edits

**Table1**
```{r Table 1}
temp <- subset(dd,!is.na(Insect))

#Run the model
lm2 <- lmer(RTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + (1|LayYear) + (1|BirdID) + (1|PlateID),
            data = temp,
            REML = FALSE)
CIs <- confint.merMod(lm2)
t1 <- data.frame(Estimate = fixef(lm2), s.e. = se.fixef(lm2), LCI = CIs[5:nrow(CIs),1], UCI = CIs[5:nrow(CIs),2])
kable(t1)

@@

```{r Table 2}
temp <- subset(dd3,!is.na(Insect))

#Run the model
lm2 <- lmer(DeltaRTL ~  LogAge + Tarsus + Helper + GroupSize + Sex + BodyMass + Insect + Density + (1|LayYear) + (1|BirdID) + (1|PlateID),
            data = temp,
            REML = FALSE)
CIs <- confint.merMod(lm2)
t1 <- data.frame(Estimate = fixef(lm2), s.e. = se.fixef(lm2), LCI = CIs[5:nrow(CIs),1], UCI = CIs[5:nrow(CIs),2])
kable(t1)

>>>>

**Table1**
```{r Table 1}

CIs <- CIwald(RTL_Full)
t1 <- data.frame(Estimate = fixef(RTL_Full), LCI = CIs[6:nrow(CIs),1], UCI = CIs[6:nrow(CIs),2])
kable(t1)

@@

```{r Table 2}

temp <- subset(dd3,!is.na(Insect))

#Run the model
CIs <- CIwald(DeltaRTL_Full)
t1 <- data.frame(Estimate = fixef(DeltaRTL_Full), LCI = CIs[4:nrow(CIs),1], UCI = CIs[4:nrow(CIs),2])
kable(t1)
